---
title: "Getting Started with normet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with normet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(normet)
```

## Introduction

`normet` provides a workflow for **meteorological normalisation** (de-weathering)
of air quality time series, together with tools for **time-series decomposition**
and **causal inference** (Synthetic Control Methods).

The typical workflow is:

1. Prepare the data (`nm_prepare_data()`).
2. Train a machine-learning model that predicts the pollutant from time and
   meteorological variables (`nm_build_model()` / `nm_train_model()`).
3. Use the model to remove the influence of meteorology, revealing the
   underlying trend (`nm_normalise()`).
4. Optionally decompose the trend further, look at rolling windows, partial
   dependence, or run a counterfactual analysis with Synthetic Control Methods.

This vignette walks through that workflow using the `my1` and `scm` datasets
bundled with the package. To keep the vignette fast to build, the model
configurations below use very small search budgets (`n_trials = 1`); for real
analyses you will normally want larger budgets (the function defaults).

```{r, eval = !requireNamespace("lightgbm", quietly = TRUE), echo = FALSE, results = "asis"}
cat("**Note:** the `lightgbm` package is not installed, so the model-training",
    "examples below are not evaluated. Install it with",
    "`install.packages(\"lightgbm\")` to run the full workflow.\n")
```

## Example data

`my1` contains real hourly NO2 observations from the UK AURN station at
London Marylebone Road (MY1, kerbside) merged with ERA5 meteorology, covering
January to August 2020 — the meteorological-normalisation case study from the
normet model-description paper. The window spans the UK COVID-19 lockdown
(from 2020-03-23), so normalisation reveals a clear emission-driven
change:

```{r}
dim(my1)
my1[1:3, c("date", "NO2", "ws", "wd", "temp", "RH", "blh", "ssrd")]
```

`scm` is a real monthly panel of observed and meteorologically normalised
(`NO2_dw`) NO2 for 104 UK
monitoring sites (2016–2021), used in the Synthetic Control Method section
below to estimate the effect of the London Ultra Low Emission Zone (ULEZ,
launched 2019-04-08) at the treated site MY1:

```{r}
dim(scm)
scm[scm$code == "MY1", ][1:3, ]
```

Two further datasets support the transport-aware normalisation case study:
`my1_pm25` (hourly PM2.5 and meteorology at MY1) and `traj_my1` (6-hourly
HYSPLIT back-trajectory features arriving at MY1), plus a two-day sample of
raw HYSPLIT tdump files in `system.file("extdata", "traj", package =
"normet")`.

## 1. Prepare and train a model

`nm_build_model()` standardises the date column, adds time features
(`date_unix`, `day_julian`, `weekday`, `hour`), splits the data into training
and testing sets, and trains a model in one step. It returns both the prepared
data (`df_prep`) and the trained model.

```{r, eval = requireNamespace("lightgbm", quietly = TRUE)}
covariates <- c("ws", "wd", "temp", "RH", "blh", "ssrd")
predictors <- c(covariates, "date_unix", "day_julian", "weekday", "hour")

model_config <- list(
  n_trials = 1, cv_folds = 2, nrounds = 15,
  num_leaves_min = 5, num_leaves_max = 15
)

build <- nm_build_model(
  my1[1:150, c("date", "NO2", covariates)],
  target = "NO2", covariates = predictors,
  model_config = model_config, seed = 42, verbose = FALSE
)

head(build$df_prep)
nm_extract_features(build$model)
```

The target column is renamed to `value` in `df_prep`, regardless of its
original name. Predictions for any data frame containing the predictor
columns can be obtained with `nm_predict()`:

```{r, eval = requireNamespace("lightgbm", quietly = TRUE)}
preds <- nm_predict(build$model, build$df_prep)
head(preds)
```

## 2. Save and reload the model

Models can be persisted with `nm_save_model()` and restored with
`nm_load_model()`. Both dispatch automatically based on the model's backend
(`lightgbm` or `h2o`).

```{r, eval = requireNamespace("lightgbm", quietly = TRUE)}
tmp <- tempfile()
nm_save_model(build$model, path = tmp, filename = "model.rds", verbose = FALSE)
model2 <- nm_load_model(path = tmp, filename = "model.rds", backend = "lightgbm", verbose = FALSE)
unlink(tmp, recursive = TRUE)

identical(nm_extract_features(build$model), nm_extract_features(model2))
```

## 3. Meteorological normalisation

`nm_normalise()` repeatedly resamples the meteorological predictors and
averages the resulting predictions, producing a meteorologically normalised
("de-weathered") series that
isolates the trend from weather-driven variability.

```{r, eval = requireNamespace("lightgbm", quietly = TRUE)}
norm <- nm_normalise(
  build$df_prep, build$model,
  resample_vars = covariates, n_samples = 5, n_cores = 1, verbose = FALSE
)
head(norm)
nm_plot_normalise(norm)
```

## 4. How many samples? Adaptive convergence

The precision of the normalised series improves with the number of
Monte-Carlo resamples, following the central-limit ~1/sqrt(n) law. Rather
than guessing `n_samples`, `nm_normalise_auto()` draws resamples in batches
and stops automatically once the *whole series* is converged: by default
(`convergence_metric = "series"`) it stops when the 95th percentile of the
per-timestamp relative standard error of the cumulative mean falls below the
tolerance for a stability streak of consecutive checks. The legacy
`"global"` criterion (batch-to-batch change of the global mean) is retained
as an option but weakens as sampling proceeds and can stop while individual
timestamps still carry sampling noise.

```{r, eval = requireNamespace("lightgbm", quietly = TRUE)}
auto <- nm_normalise_auto(
  build$df_prep, build$model, resample_vars = covariates,
  batch_size = 2, max_samples = 10, convergence_tol = 0.2,
  n_cores = 1, verbose = FALSE, return_history = TRUE
)
auto$best_n
auto$history[, c("n", "metric", "stable_count")]
```

(The tolerance here is deliberately loose so the vignette builds quickly;
the default — 3% on the per-timestamp relative standard error — is
appropriate for real analyses.)

## 5. Decomposition

`nm_decompose()` separates the series into components attributable to
different drivers. With `method = "emission"`, it isolates the contribution
of the time-based features (trend, seasonality, weekday, hour) from the
meteorology-driven "noise":

```{r, eval = requireNamespace("lightgbm", quietly = TRUE)}
decomp <- nm_decompose(
  method = "emission", df = build$df_prep, model = build$model,
  covariates = predictors, n_samples = 2, n_cores = 1, verbose = FALSE
)
head(decomp)
```

## 6. Rolling normalisation

For trend analysis, `nm_rolling()` applies the normalisation within a moving
window, producing one normalised value per window for each timestamp:

```{r, eval = requireNamespace("lightgbm", quietly = TRUE)}
roll <- nm_rolling(
  build$df_prep, model = build$model, covariates = predictors,
  resample_vars = covariates, window_days = 4, rolling_every = 8,
  n_samples = 1, n_cores = 1, verbose = FALSE
)
head(roll)
```

## 7. Transport-aware normalisation

Standard normalisation resamples only **local** meteorology, but pollutants
like PM2.5 at a kerbside site are also shaped by **long-range transport**.
Augmenting the predictors with HYSPLIT back-trajectory features lets the
normalisation marginalise over transport too; the difference between the
local-only and transport-aware normalised series is the transport
contribution. This is the paper's COVID-19 PM2.5 case study, supported by
the bundled `my1_pm25` (hourly PM2.5 + ERA5 meteorology) and `traj_my1`
(6-hourly trajectory-derived source-sector features) datasets.

Raw HYSPLIT endpoint files can be parsed with `nm_read_trajectory_tdump()`
(a two-day sample ships with the package), and `traj_my1` was built from
such files with `nm_build_trajectory_features()`:

```{r}
tdump <- system.file("extdata", "traj", "tdump_2020010100", package = "normet")
head(nm_read_trajectory_tdump(tdump), 3)
traj_my1[1:3, 1:6]
```

Join the 6-hourly trajectory features onto the hourly PM2.5 panel and train
two models — local meteorology only, and transport-aware — then normalise
with each:

```{r, eval = requireNamespace("lightgbm", quietly = TRUE)}
met_cols <- c("ws_era5", "wd_era5", "temp_era5", "blh", "sp", "ssrd", "tcc", "tp")
traj_cols <- grep("^traj_", names(traj_my1), value = TRUE)

pm <- my1_pm25[!is.na(my1_pm25$pm25), c("date", "pm25", met_cols)]
pm$date6 <- as.POSIXct(round(as.numeric(pm$date) / 21600) * 21600,
                       origin = "1970-01-01", tz = "UTC")
traj6 <- traj_my1
names(traj6)[names(traj6) == "date"] <- "date6"
pm <- merge(pm, traj6[, c("date6", traj_cols)], by = "date6")
pm <- pm[order(pm$date), c("date", "pm25", met_cols, traj_cols)][1:200, ]

preds_local <- c(met_cols, "date_unix", "day_julian", "weekday", "hour")
preds_transport <- c(met_cols, traj_cols, "date_unix", "day_julian", "weekday", "hour")

b1 <- nm_build_model(pm[, c("date", "pm25", met_cols)], target = "pm25",
                     covariates = preds_local,
                     model_config = model_config, seed = 42, verbose = FALSE)
dw_local <- nm_normalise(b1$df_prep, b1$model, resample_vars = met_cols,
                         n_samples = 5, n_cores = 1, verbose = FALSE)

b2 <- nm_build_model(pm, target = "pm25", covariates = preds_transport,
                     model_config = model_config, seed = 42, verbose = FALSE)
dw_transport <- nm_normalise(b2$df_prep, b2$model,
                             resample_vars = c(met_cols, traj_cols),
                             n_samples = 5, n_cores = 1, verbose = FALSE)

transport_contribution <- dw_local$normalised - dw_transport$normalised
summary(transport_contribution)
```

Positive values flag hours where a local-only normalisation would attribute
transport-driven concentration to local emissions. (With the tiny training
subset and budgets used here this is illustrative only — see the paper for
the full eight-month analysis.)

## 8. Partial dependence

`nm_pdp()` computes partial dependence values for one or more predictors,
showing how the model's predictions vary with that predictor while
marginalising over the others:

```{r, eval = requireNamespace("lightgbm", quietly = TRUE)}
pdp <- nm_pdp(build$df_prep, build$model, var_list = c("temp"), n_cores = 1, verbose = FALSE)
head(pdp)
nm_plot_pdp(pdp)
```

## 9. End-to-end pipeline

`nm_do_all()` chains data preparation, model training, and normalisation in a
single call, returning the normalised result, the trained model, and the
prepared data:

```{r, eval = requireNamespace("lightgbm", quietly = TRUE)}
result <- nm_do_all(
  my1[1:150, c("date", "NO2", covariates)],
  target = "NO2", covariates = predictors,
  model_config = model_config, n_samples = 5, seed = 42,
  n_cores = 1, memory_save = FALSE, verbose = FALSE
)
names(result)
head(result$res)
```

## 10. Causal inference: Synthetic Control Method

`nm_run_scm()` provides a unified interface to several Synthetic Control
Method variants (`nm_scm()` is the augmented SCM implementation it dispatches
to by default). Given a panel of one treated unit and several donor units, it
estimates a synthetic counterfactual for the treated unit and the resulting
treatment effect:

Following the paper's primary ULEZ design, we take the pre-COVID window
(2017-01 to 2020-02), use traffic-type sites as the donor pool, and keep
donors with complete coverage so the donor matrix is fully defined:

```{r}
win <- scm[scm$date >= as.Date("2017-01-01") & scm$date < as.Date("2020-03-01"), ]
traffic <- win[grepl("Traffic", win$type), ]
n_months <- length(unique(traffic$date))
cov <- tapply(!is.na(traffic$NO2_dw), traffic$code, sum)
panel <- traffic[traffic$code %in% names(cov)[cov == n_months], ]

res <- nm_run_scm(
  df = panel, date_col = "date", unit_col = "code", outcome_col = "NO2_dw",
  treated_unit = "MY1", cutoff_date = "2019-04-01",
  scm_backend = "scm", verbose = FALSE
)
head(res)
nm_plot_time_series(res, target = "effect", title = "ULEZ effect on normalised NO2 at MY1")
```

## Next steps

This vignette covers the core workflow. See the function reference
(`?nm_normalise`, `?nm_decompose`, `?nm_run_scm`, etc.) for additional
options, including the H2O AutoML backend (`backend = "h2o"`), other SCM
variants (`scm_backend = "mlscm"`, `"abadie"`, `"did"`, `"mcnnm"`), and
significance/diagnostic tools for causal estimates.
