---
title: "Getting Started with normet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with normet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(normet)
```

## Introduction

`normet` provides a workflow for **meteorological normalisation** (de-weathering)
of air quality time series, together with tools for **time-series decomposition**
and **causal inference** (Synthetic Control Methods).

The typical workflow is:

1. Prepare the data (`nm_prepare_data()`).
2. Train a machine-learning model that predicts the pollutant from time and
   meteorological variables (`nm_build_model()` / `nm_train_model()`).
3. Use the model to remove the influence of meteorology, revealing the
   underlying trend (`nm_normalise()`).
4. Optionally decompose the trend further, look at rolling windows, partial
   dependence, or run a counterfactual analysis with Synthetic Control Methods.

This vignette walks through that workflow using the `my1` and `scm` datasets
bundled with the package. To keep the vignette fast to build, the model
configurations below use very small search budgets (`n_trials = 1`); for real
analyses you will normally want larger budgets (the function defaults).

```{r, eval = !requireNamespace("lightgbm", quietly = TRUE), echo = FALSE, results = "asis"}
cat("**Note:** the `lightgbm` package is not installed, so the model-training",
    "examples below are not evaluated. Install it with",
    "`install.packages(\"lightgbm\")` to run the full workflow.\n")
```

## Example data

`my1` is a synthetic hourly air-quality and meteorology dataset modelled on the
MY1 (London Marylebone Road) monitoring station, covering the full year 2020:

```{r}
dim(my1)
my1[1:3, c("date", "NO2", "u10", "v10", "d2m", "t2m", "blh", "ssrd")]
```

`scm` is a synthetic weekly panel of pollutant concentrations for one treated
unit ("2+26 cities") and 37 donor cities, used in the Synthetic Control Method
section below:

```{r}
dim(scm)
scm[1:3, c("date", "ID", "NO2")]
```

## 1. Prepare and train a model

`nm_build_model()` standardises the date column, adds time features
(`date_unix`, `day_julian`, `weekday`, `hour`), splits the data into training
and testing sets, and trains a model in one step. It returns both the prepared
data (`df_prep`) and the trained model.

```{r, eval = requireNamespace("lightgbm", quietly = TRUE)}
covariates <- c("u10", "v10", "d2m", "t2m", "blh", "ssrd")
predictors <- c(covariates, "date_unix", "day_julian", "weekday", "hour")

model_config <- list(
  n_trials = 1, cv_folds = 2, nrounds = 15,
  num_leaves_min = 5, num_leaves_max = 15
)

build <- nm_build_model(
  my1[1:150, c("date", "NO2", covariates)],
  value = "NO2", predictors = predictors,
  model_config = model_config, seed = 42, verbose = FALSE
)

head(build$df_prep)
nm_extract_features(build$model)
```

The target column is renamed to `value` in `df_prep`, regardless of its
original name. Predictions for any data frame containing the predictor
columns can be obtained with `nm_predict()`:

```{r, eval = requireNamespace("lightgbm", quietly = TRUE)}
preds <- nm_predict(build$model, build$df_prep)
head(preds)
```

## 2. Save and reload the model

Models can be persisted with `nm_save_model()` and restored with
`nm_load_model()`. Both dispatch automatically based on the model's backend
(`lightgbm` or `h2o`).

```{r, eval = requireNamespace("lightgbm", quietly = TRUE)}
tmp <- tempfile()
nm_save_model(build$model, path = tmp, filename = "model.rds", verbose = FALSE)
model2 <- nm_load_model(path = tmp, filename = "model.rds", backend = "lightgbm", verbose = FALSE)
unlink(tmp, recursive = TRUE)

identical(nm_extract_features(build$model), nm_extract_features(model2))
```

## 3. Weather normalisation

`nm_normalise()` repeatedly resamples the meteorological predictors and
averages the resulting predictions, producing a "de-weathered" series that
isolates the trend from weather-driven variability.

```{r, eval = requireNamespace("lightgbm", quietly = TRUE)}
norm <- nm_normalise(
  build$df_prep, build$model,
  resample_vars = covariates, n_samples = 5, n_cores = 1, verbose = FALSE
)
head(norm)
nm_plot_normalise(norm)
```

## 4. Decomposition

`nm_decompose()` separates the series into components attributable to
different drivers. With `method = "emission"`, it isolates the contribution
of the time-based features (trend, seasonality, weekday, hour) from the
meteorology-driven "noise":

```{r, eval = requireNamespace("lightgbm", quietly = TRUE)}
decomp <- nm_decompose(
  method = "emission", df = build$df_prep, model = build$model,
  predictors = predictors, n_samples = 2, n_cores = 1, verbose = FALSE
)
head(decomp)
```

## 5. Rolling normalisation

For trend analysis, `nm_rolling()` applies the normalisation within a moving
window, producing one normalised value per window for each timestamp:

```{r, eval = requireNamespace("lightgbm", quietly = TRUE)}
roll <- nm_rolling(
  build$df_prep, model = build$model, predictors = predictors,
  resample_vars = covariates, window_days = 4, rolling_every = 8,
  n_samples = 1, n_cores = 1, verbose = FALSE
)
head(roll)
```

## 6. Partial dependence

`nm_pdp()` computes partial dependence values for one or more predictors,
showing how the model's predictions vary with that predictor while
marginalising over the others:

```{r, eval = requireNamespace("lightgbm", quietly = TRUE)}
pdp <- nm_pdp(build$df_prep, build$model, var_list = c("t2m"), n_cores = 1, verbose = FALSE)
head(pdp)
nm_plot_pdp(pdp)
```

## 7. End-to-end pipeline

`nm_do_all()` chains data preparation, model training, and normalisation in a
single call, returning the normalised result, the trained model, and the
prepared data:

```{r, eval = requireNamespace("lightgbm", quietly = TRUE)}
result <- nm_do_all(
  my1[1:150, c("date", "NO2", covariates)],
  value = "NO2", predictors = predictors,
  model_config = model_config, n_samples = 5, seed = 42,
  n_cores = 1, memory_save = FALSE, verbose = FALSE
)
names(result)
head(result$res)
```

## 8. Causal inference: Synthetic Control Method

`nm_run_scm()` provides a unified interface to several Synthetic Control
Method variants (`nm_scm()` is the augmented SCM implementation it dispatches
to by default). Given a panel of one treated unit and several donor units, it
estimates a synthetic counterfactual for the treated unit and the resulting
treatment effect:

```{r}
res <- nm_run_scm(
  df = scm, date_col = "date", unit_col = "ID", outcome_col = "NO2",
  treated_unit = "2+26 cities", cutoff_date = "2015-12-01",
  scm_backend = "scm", verbose = FALSE
)
head(res)
nm_plot_time_series(res, value = "effect", title = "SCM effect on NO2")
```

## Next steps

This vignette covers the core workflow. See the function reference
(`?nm_normalise`, `?nm_decompose`, `?nm_run_scm`, etc.) for additional
options, including the H2O AutoML backend (`backend = "h2o"`), other SCM
variants (`scm_backend = "mlscm"`, `"abadie"`, `"did"`, `"mcnnm"`), and
significance/diagnostic tools for causal estimates.
